### 2017-06-14 paper/prelims.tex
Streamlined the current vesrion of the paper a bit, added note on integrability.

\vspace{-1em}
\subsection{Basic Notions, Linear Predicates, Valuations}
\vspace{-0.5em}
For a set $A$ we denote by $|A|$ the cardinality of $A$. We denote by $\Nset$,
$\Nset_0$, $\Zset$, and $\Rset$ the sets of all positive integers, non-negative
integers, integers, and real numbers, respectively. We assume basic knowledge 
of matrix calculus.
We use boldface notation for
vectors, e.g. $\vec{x}$, $\vec{y}$, etc., and we denote an $i$-th component of a
vector $\vec{x}$ by $\vec{x}[i]$. For the purpose of matrix calculations we 
assume that (non-transposed) vectors are row vectors. If $\vec{v},\vec{v}'$ are 
$n$ and $m$ dimensional vectors, respectively, then $(\vec{v},\vec{v}')$ is an 
$(n+m)$-dimensional vector obtained by ``concatenation'' of $\vec{v}$ and 
$\vec{v}'$. We identify 1-dimensional vectors with numbers. For an 
$n$-dimensional vector 
$\vec{x}$, index $1 \leq i\leq n$, and number $a$ we denote by $\vec{x}(i\leftarrow a)$ 
a 
vector $\vec{y}$ such that $\vec{y}[i]=a$ and $\vec{y}[j]=\vec{x}[j]$ for all 
$1\leq j \leq n$, $j\neq i$.
For comparison of vectors (e.g. as in $\vec{x}\leq \vec{y}$), we consider componentwise 
comparison. 
For comparing functions $f,g$ with the same domains, we write $f\leq g$ 
if $f(x)\leq g(x)$ for all $x$ in the domain.

@@

\smallskip\noindent{\em Expressions.}
An \emph{expression} over the set of variables $\{x_1,\dots,x_n\}$ 
is an expression in the standard programming-language sense, i.e. a formula 
built in finite number of steps from constants, variables $x_1,\dots,x_n$, and 
numerical operators from some fixed finite set. Each expression $E$ over 
$\{x_1,\dots,x_n\}$ 
determines a function which for each $m$-dimensional vector $\vec{x}$, where 
$m\geq n$,  
returns a number resulting from substituting each $x_i$ in $E$ by $\vec{x}[i]$. 
Slightly abusing our notation, we denote this function also by $E$ and the 
value of this function on argument $\vec{x}$ by $E(\vec{x})$. 
%An expression $E$ 
%is \emph{measurable} if the function $E(\vec{x})$ is Borel 
%measurable~\cite{xxx}.  In particular, expressions built using the standard 
%arithmetic operators  are all measurable.

\smallskip\noindent{\em Affine Expressions.}
An \emph{affine expression} over the set of variables $\{x_1,\dots,x_n\}$ is an 
expression of the form $d+\sum_{i=1}^{n}a_i
x_i$, where  $d,a_1,\dots,a_n$ are real-valued
constants.  A function of the form $E(\vec{x})$ for some affine expression $E$ 
is called affine.
%Following the terminology of~\cite{DBLP:conf/sas/KatoenMMM10} we fix the
%following nomenclature:

\smallskip\noindent{\em Linear constraint, assertion, predicates.}
We use the following nomenclature:
\begin{compactitem}
\item {\em Linear Constraint.} A \emph{linear constraint} is a formula of the
form $\psi$ or $\neg\psi$, where  $\psi$ is a non-strict inequality
between affine expressions.
\item {\em Linear Assertion.} A \emph{linear assertion} is a finite conjunction
of linear constraints.
\item {\em Propositionally Linear Predicate.}
A  \emph{propositionally linear predicate} (PLP) is a finite disjunction of
linear assertions.
\end{compactitem}

\noindent{\em Arity and satisfaction of PLP.}
%The \emph{arity} of a PLP is the number of distinct variables appearing in it. 
For a PLP $\varphi$ we denote by $\vars(\varphi)$ the set of all variables that 
appear in $\varphi$. As noted above, we stipulate that $\vars(\varphi)=\{x_1,\dots,x_{n(\varphi)}\}$ for some $n(\varphi)\in \Nset$. A vector $\vec{x}$ of dimension $m\geq n(\varphi)$ \emph{satisfies} $\varphi$, we write 
$\vec{x}\models\phi$, if 
 the arithmetic formula obtained by 
substituting each occurrence of a variable $x_i$ in $\varphi$ by $\vec{x}[i]$ is 
valid. We denote $\sat{\phi} = \{\vec{x}\in \Rset^m\mid m\geq n(\varphi) \wedge \vec{x} \models\phi\}$ 
and $\exsat{\phi} = \sat{\phi}
\cap \Rset^{n(\varphi)}$.

@@

\noindent{\em The Syntax.}
We consider the standard syntax for probabilistic programs,
which encompasses basic programming mechanisms such as assignment statement 
(indicated by `:='), while-loop, if-branch. We also consider basic probabilistic mechanisms 
such as probabilistic branch (indicated by `prob') and random sampling (e.g. 
$x:=\textbf{sample(}\mathrm{Uniform}[-2,1]\textbf{)}$ assigns to $x$ a random 
number 
uniformly sampled from interval $[-2,1]$). We also allow constructs for 
(demonic) non-determinism, in particular 
non-deterministic branching indicated by `\textbf{if }$\star$ \textbf{then...}' construct and non-deterministic assignment.  
Variables (or identifiers) of a probabilistic program are of \emph{real} type, i.e., 
values of the variables are real numbers. 
We also assume that assume that each \APP{} $\program$ is preceded by an 
initialization preamble 
in 
which 
each variable appearing in $\program$ is assigned some concrete number.
the program variables are initialized. We do not a priori (\textbf{PETR: 
elaborate one xpressions})

@@

\vspace{-1em}
\subsection{Semantics of Affine Probabilistic Programs}\label{subsec:semantics}
\vspace{-0.5em}

We now formally define the semantics of \APP's.
In order to do this, we first recall some fundamental concepts from probability
theory.

@@

\smallskip
\begin{definition}
\label{def:stochgame}
A \emph{probabilistic control flow graph (pCFG)} is a tuple
$\pCFG=(\locs,\pvars,\locinit,\vecinit,\transitions,\updates,\probdist,\guards)$,
where
\begin{compactitem}
\item $\locs$ is a finite set of \emph{locations} partitioned into four 
pairwise
disjoint subsets  $\locsNB$, $\locsPB$, $\locs_D$, and $\locs_A$ of 
non-deterministic branching, 
probabilistic branching, deterministic, and assignment locations;
\item $\pvars=\{x_1,\dots,x_{|\pvars|}\}$ is a finite set of \emph{program 
variables} (note that $\pvars \subseteq \vars$) ;
\item $\locinit$ is an initial location and $\vecinit$ is an initial 
\emph{assignment vector};
% POKRACOVAT ODSUD
\item $\transitions\subseteq\locs\times\locs$ is a transition relation'
%whose members are tuples of
%the form $(\ell,i,\up,\ell')$, where $\ell$ and $\ell'$ are source and target
%program locations, respectively, $1\leq i \leq |\pvars|$ is a \emph{target 
%variable index}, and $u$ 
%is an 
%\emph{update element}, which can 
%be one of the following mathematical objects: %\begin{itemize}
%%\item 
%(a)~an affine function $u\colon \Rset^{|\pvars|}\rightarrow \Rset$;
%%\item 
%(b)~a distribution $d\in \mathcal{D}$; or
%%%\item 
%(c)~a set $R\subseteq \Rset$.
%\end{itemize}
\item $\updates$ is a function assigning to each transition outgoing from an 
assignment location a tuple $(i,\up)$, where $1\leq i \leq |\pvars|$ is a 
\emph{target 
	variable index} and $\up$ 
is an 
\emph{update element}, which can 
be one of the following mathematical objects: %\begin{itemize}
%\item 
(a)~an affine function $u\colon \Rset^{|\pvars|}\rightarrow \Rset$;
%\item 
(b)~a distribution $d\in \mathcal{D}$; or
%\item 
(c)~a set $R\subseteq \Rset$.
%\end{itemize}
\item $\probdist=\{\prob_{\ell}\}_{\ell \in \locsPB}$ is a collection of
probability distributions, where each $\prob_{\ell}$ is a discrete probability
distribution on the set of all transitions outgoing from~$\ell$;
\item $\guards$ is a function assigning a propositionally linear predicate
(a \emph{guard}) over $\pvars$ to each transition outgoing from a deterministic 
location.
\end{compactitem}

@@

\smallskip\noindent{\em Executions and reachable configurations.}
We say that a configuration $(\loc',\vec{x}')$ is a \emph{successor} of a 
configuration $(\loc,\vec{x})$ if 
%A \emph{finite path} (or
%\emph{execution fragment}) of length $k$ in $\pCFG$ is a finite sequence of
%configurations $(\ell_0,\vec{x}_0)\cdots(\ell_k,\vec{x}_k)$ such that 
%%$(\loc_0,\vec{x}_0)=(\locinit,\vecinit)$ and 
%for each
%$0 \leq i < k$ 
there is a transition  $\tau=(\loc,\loc')$ enabled in
$(\loc,\vec{x})$ 
%such that $\vec{x}_{i+1}=\vec{x}_i(j\leftarrow a)$ where $a$ 
%satisfies one of the following:
and $\vec{x}'$ satisfies the following:
\begin{compactitem}
\item if $\loc$ is not an assignment location, then $\vec{x}'=\vec{x}$;
\item if $\loc$ is an assignment location with $\updates(\tau)=(j,\up)$, then 
$\vec{x}_{i+1}=\vec{x}_i(j\leftarrow a)$ where $a$ satisfies one of the 
following 
depending on the type of $\up$:
\begin{compactitem}
\item if $\up$ is an affine function, then $a=\up(\vec{x})$;
\item if $\up$ is an integrable\footnote{A distribution on some numerical 
domain 
is integrable if its expected value exists and is finite. In particular, each 
Dirac distribution is integrable.} distribution $d$, then 
$a\in \support(d)$;
\item if $\up$ is a set, then $a$ is some element of $\up$.
\end{compactitem}
\end{compactitem}
A \emph{finite path} (or
\emph{execution fragment}) of length $k$ in $\pCFG$ is a finite sequence of
configurations $(\ell_0,\vec{x}_0)\cdots(\ell_k,\vec{x}_k)$ such that 
%$(\loc_0,\vec{x}_0)=(\locinit,\vecinit)$ and 
for each
$0 \leq i < k$ the configuration $(\loc_{i+1},\vec{x}_{i+1})$ is a successor of 
$(\loc_i,\vec{x}_i)$.
A \emph{run} (or \emph{execution}) in
$\pCFG$ is an infinite sequence of configurations whose every finite
prefix is a finite path.
A configuration $(\loc,\vec{x})$ is {\em reachable} from the initial 
configuration
$(\locinit,\vecinit)$
if there is a finite path starting in $(\locinit,\vecinit)$ that ends in
$(\loc,\vec{x})$. We denote by $\Conf_\pCFG,\Fpath_\pCFG$ and $\Run_\pCFG$ the 
sets of all configurations, finite paths and runs in $\pCFG$, respectively, 
dropping the index $\pCFG$ when known from the context.

@@

\smallskip\noindent{\em Termination, reachability, and termination time.}
In the following, consider an \APP{} $P$ and its associated pCFG $\pCFG_P$. 
We say that a run $\run$ of $\pCFG_P$ \emph{reaches} a set of configurations 
$\confset$ if it  contains a configuration from $\confset$. A run 
\emph{terminates} if it reaches a 
configuration whose first component (i.e. a location of $\pCFG_P$) is the 
location
$\loc_{P}^{\lout}$ corresponding to the value of the program counter after 
executing $P$. 
To each set of configurations $\confset$ we can assign a random variable 
$\treach{\confset}$ 
such that for each run $\run$ the value $\treach{\confset}(\run)$ represents 
the first 
point in time when 
the current configuration on $\run$ is in $\confset$. If a run $\run$ does 
\emph{not} 
reach a set $\confset$, then 
$\treach{\confset}(\run)=\infty$. We call $\treach{\confset}$ the 
\emph{reachability time} of $\confset$. In particular, if $\confset$ is the set 
of all configurations $(\loc,\vec{x})$ such that $\loc=\loc^\lout_P$ (the 
terminal location of $\pCFG_P$), then $\treach{\confset}$ is 
called a \emph{termination time}, as it returns the number of steps after which 
$\run$ terminates. Since termination time is an important concept on its own, 
we use a special notation $\ttime$ for it.
%\begin{example}
%Consider the program $P$ such that \hspace{-0.25cm}\newsavebox{\figa}
%\lstset{language=affprob}
%\lstset{tabsize=1}
%\begin{lrbox}{\figa}
%\begin{lstlisting}[mathescape]
%$P=$x:=2;y:=1
%\end{lstlisting}
%\end{lrbox}
%\usebox{\figa}. The associated pCFG has three locations: initial location 
%$\loc_{P}^{\lin}$, terminal location $\loc_{P}^{\lout}$, and one 
%``intermediate'' location, say $\loc$. In the pCFG consider a run 
%$\run=(\loc_{P}^{\lin},(0,0))(\loc,(2,0))(\loc_{P}^{\lout},(2,1))(\loc_{P}^{\lout},(2,1))\cdots$.
% Then $\ttime(\run)=2$. If we denote by $C$ the set of configurations 
% $\{(\loc',(x,y))\mid \loc' \in\{\loc_{P}^{\lin},\loc_{P}^{\lout},\loc\} \text{ 
% and } x\geq 2\}$, then $\treach{C}(\run)=1$.
%\end{example}
Since a probabilistic program may exhibit more than one run, we are interested 
in probabilities of runs that terminate or reach some set of configurations. 
This gives rise to the following fundamental computational problems regarding 
termination:

>>>>

\vspace{-1em}
\subsection{Basic Notions}
\vspace{-0.5em}
For a set $A$ we denote by $|A|$ the cardinality of $A$. We denote by $\Nset$,
$\Nset_0$, $\Zset$, and $\Rset$ the sets of all positive integers, non-negative
integers, integers, and real numbers, respectively. We assume basic knowledge 
of matrix calculus.
We use boldface notation for
vectors, e.g. $\vec{x}$, $\vec{y}$, etc., and we denote an $i$-th component of a
vector $\vec{x}$ by $\vec{x}[i]$. For the purpose of matrix calculations we 
assume that (non-transposed) vectors are row vectors. If $\vec{v},\vec{v}'$ are 
$n$ and $m$ dimensional vectors, respectively, then $(\vec{v},\vec{v}')$ is an 
$(n+m)$-dimensional vector obtained by ``concatenation'' of $\vec{v}$ and 
$\vec{v}'$. We identify 1-dimensional vectors with numbers. For an 
$n$-dimensional vector 
$\vec{x}$, index $1 \leq i\leq n$, and number $a$ we denote by $\vec{x}(i\leftarrow a)$ 
a 
vector $\vec{y}$ such that $\vec{y}[i]=a$ and $\vec{y}[j]=\vec{x}[j]$ for all 
$1\leq j \leq n$, $j\neq i$.
For comparison of vectors (e.g. as in $\vec{x}\leq \vec{y}$), we consider componentwise 
comparison. 
For comparing functions $f,g$ with the same domains, we write $f\leq g$ 
if $f(x)\leq g(x)$ for all $x$ in the domain.

@@

%\smallskip\noindent{\em Expressions.}
%An \emph{expression} over the set of variables $\{x_1,\dots,x_n\}$ 
%is an expression in the standard programming-language sense, i.e. a formula 
%built in finite number of steps from constants, variables $x_1,\dots,x_n$, and 
%numerical operators from some fixed finite set. Each expression $E$ over 
%$\{x_1,\dots,x_n\}$ 
%determines a function which for each $m$-dimensional vector $\vec{x}$, where 
%$m\geq n$,  
%returns a number resulting from substituting each $x_i$ in $E$ by $\vec{x}[i]$. 
%Slightly abusing our notation, we denote this function also by $E$ and the 
%value of this function on argument $\vec{x}$ by $E(\vec{x})$. 
%%An expression $E$ 
%%is \emph{measurable} if the function $E(\vec{x})$ is Borel 
%%measurable~\cite{xxx}.  In particular, expressions built using the standard 
%%arithmetic operators  are all measurable.
%
%\smallskip\noindent{\em Affine Expressions.}
%An \emph{affine expression} over the set of variables $\{x_1,\dots,x_n\}$ is an 
%expression of the form $d+\sum_{i=1}^{n}a_i
%x_i$, where  $d,a_1,\dots,a_n$ are real-valued
%constants.  A function of the form $E(\vec{x})$ for some affine expression $E$ 
%is called affine.
%%Following the terminology of~\cite{DBLP:conf/sas/KatoenMMM10} we fix the
%%following nomenclature
%
%\smallskip\noindent{\em Predicates}
%A \emph{predicate} is a logical formula obtained by a finite number of applications of conjunction, disjunction and negation operations on \emph{atomic predicates} of the form $E(\vec{x})\leq E'(\vec{y})$, where $E$, $E'$ are expressions.
%
%\smallskip\noindent{\em Linear constraints, assertions, predicates.}
%In the case of predicates involving only linear expression we use the following nomenclature:
%\begin{compactitem}
%\item {\em Linear Constraint.} A \emph{linear constraint} is a formula of the
%form $\psi$ or $\neg\psi$, where  $\psi$ is a non-strict inequality
%between affine expressions.
%\item {\em Linear Assertion.} A \emph{linear assertion} is a finite conjunction
%of linear constraints.
%\item {\em Propositionally Linear Predicate.}
%A  \emph{propositionally linear predicate} (PLP) is a finite disjunction of
%linear assertions.
%\end{compactitem}
%
%\noindent{\em Arity and satisfaction of PLP.}
%%The \emph{arity} of a PLP is the number of distinct variables appearing in it. 
%For a PLP $\varphi$ we denote by $\vars(\varphi)$ the set of all variables that 
%appear in $\varphi$. As noted above, we stipulate that $\vars(\varphi)=\{x_1,\dots,x_{n(\varphi)}\}$ for some $n(\varphi)\in \Nset$. A vector $\vec{x}$ of dimension $m\geq n(\varphi)$ \emph{satisfies} $\varphi$, we write 
%$\vec{x}\models\phi$, if 
% the arithmetic formula obtained by 
%substituting each occurrence of a variable $x_i$ in $\varphi$ by $\vec{x}[i]$ is 
%valid. We denote $\sat{\phi} = \{\vec{x}\in \Rset^m\mid m\geq n(\varphi) \wedge \vec{x} \models\phi\}$ 
%and $\exsat{\phi} = \sat{\phi}
%\cap \Rset^{n(\varphi)}$.

@@

\smallskip\noindent{\em Expressions.}
An \emph{expression} over the set of variables $\{x_1,\dots,x_n\}$ 
is an expression in the standard programming-language sense, i.e. a formula 
built in finite number of steps from constants, variables $x_1,\dots,x_n$, and 
numerical operators from some fixed finite set. Each expression $E$ over 
$\{x_1,\dots,x_n\}$ 
determines a function which for each $m$-dimensional vector $\vec{x}$, where 
$m\geq n$,  
returns a number resulting from substituting each $x_i$ in $E$ by $\vec{x}[i]$. 
Slightly abusing our notation, we denote this function also by $E$ and the 
value of this function on argument $\vec{x}$ by $E(\vec{x})$. We do not a priori fix a concrete set of operators that can be used to form expressions. However, in order to ensure that semantics of probabilistic programs with real-valued variables is defined correctly, we impose the following two conditions on the set of expressions used in each program:
\begin{enumerate}
\item
For each expression $E$ over variables $\{x_1,\dots,x_n\}$ and each $n$-dimensional vector $\vec{x}$ the value $E(\vec{x})$ is well defined.\footnote{Our results can be easily extended to programs where encountering an expression of undefined value, such as division by zero, triggers an exception which terminates the program, but we abstract away from such details for the sake of clarity.}
\item
The function defined by each expression $E$ is Borel-measurable (for definition of Borel-measura\-bility, see, e.g.~\cite{Billingsley:book}).
\end{enumerate}

From measure theory it is known that these conditions hold in particular for programs where expressions are build using the standard arithmetic operators of addition, subtraction, multiplication, and division (provided that expressions evaluating to zero are not allowed as divisors).

%An expression $E$ 
%is \emph{measurable} if the function $E(\vec{x})$ is Borel 
%measurable~\cite{xxx}.  In particular, expressions built using the standard 
%arithmetic operators  are all measurable.

\smallskip\noindent{\em Affine Expressions.}
An \emph{affine expression} over the set of variables $\{x_1,\dots,x_n\}$ is an 
expression of the form $d+\sum_{i=1}^{n}a_i
x_i$, where  $d,a_1,\dots,a_n$ are real-valued
constants.  A function of the form $E(\vec{x})$ for some affine expression $E$ 
is called affine. As noted above
%Following the terminology of~\cite{DBLP:conf/sas/KatoenMMM10} we fix the
%following nomenclature

\smallskip\noindent{\em Predicates}
A \emph{predicate} is a logical formula obtained by a finite number of applications of conjunction, disjunction and negation operations on \emph{atomic predicates} of the form $E(\vec{x})\leq E'(\vec{y})$, where $E$, $E'$ are expressions.

\smallskip\noindent{\em Linear constraints, assertions, predicates.}
In the case of predicates involving only linear expression we use the following nomenclature:
\begin{compactitem}
	\item {\em Linear Constraint.} A \emph{linear constraint} is a formula of the
	form $\psi$ or $\neg\psi$, where  $\psi$ is a non-strict inequality
	between affine expressions.
	\item {\em Linear Assertion.} A \emph{linear assertion} is a finite conjunction
	of linear constraints.
	\item {\em Propositionally Linear Predicate.}
	A  \emph{propositionally linear predicate} (PLP) is a finite disjunction of
	linear assertions.
\end{compactitem}

\textbf{[PETR: SAY SOMETHING ABOUT PREDICATE SATISFACTION.]}
\textbf{[PETR: EXPLAIN BOREL-MEASURABILITY IN MORE DETAIL?]}

%\noindent{\em Arity and satisfaction of PLP.}
%%The \emph{arity} of a PLP is the number of distinct variables appearing in it. 
%For a PLP $\varphi$ we denote by $\vars(\varphi)$ the set of all variables that 
%appear in $\varphi$. As noted above, we stipulate that $\vars(\varphi)=\{x_1,\dots,x_{n(\varphi)}\}$ for some $n(\varphi)\in \Nset$. A vector $\vec{x}$ of dimension $m\geq n(\varphi)$ \emph{satisfies} $\varphi$, we write 
%$\vec{x}\models\phi$, if 
%the arithmetic formula obtained by 
%substituting each occurrence of a variable $x_i$ in $\varphi$ by $\vec{x}[i]$ is 
%valid. We denote $\sat{\phi} = \{\vec{x}\in \Rset^m\mid m\geq n(\varphi) \wedge \vec{x} \models\phi\}$ 
%and $\exsat{\phi} = \sat{\phi}
%\cap \Rset^{n(\varphi)}$.

\noindent{\em The Syntax of Probabilistic Programs {(\PP s)}.}
We consider the standard syntax for probabilistic programs,
which encompasses basic programming mechanisms such as assignment statement 
(indicated by `:='), while-loop, if-branch. Expressions appear on right-hand sides of assignments, and predicates act as loop guards and conditions in if-then-else statements. We also consider basic probabilistic mechanisms 
such as probabilistic branch (indicated by `prob') and random sampling (e.g. 
$x:=\textbf{sample(}\mathrm{Uniform}[-2,1]\textbf{)}$ assigns to $x$ a random 
number 
uniformly sampled from interval $[-2,1]$). We also allow constructs for 
(demonic) non-determinism, in particular 
non-deterministic branching indicated by `\textbf{if }$\star$ \textbf{then...}' construct and non-deterministic assignment.  
Variables (or identifiers) of a probabilistic program are of \emph{real} type, i.e., 
values of the variables are real numbers. 
We also assume that assume that each \PP{} $\program$ is preceded by an 
initialization preamble 
in 
which 
each variable appearing in $\program$ is assigned some concrete number.
the program variables are initialized.

@@

\vspace{-1em}
\subsection{Semantics of Probabilistic Programs}\label{subsec:semantics}
\vspace{-0.5em}

We now formally define the semantics of \PP's.
In order to do this, we first recall some fundamental concepts from probability
theory.

@@

\smallskip
\begin{definition}
\label{def:stochgame}
A \emph{probabilistic control flow graph (pCFG)} is a tuple
$\pCFG=(\locs,\pvars,\locinit,\vecinit,\transitions,\updates,\probdist,\guards)$,
where
\begin{compactitem}
\item $\locs$ is a finite set of \emph{locations} partitioned into four 
pairwise
disjoint subsets  $\locsNB$, $\locsPB$, $\locs_D$, and $\locs_A$ of 
non-deterministic branching, 
probabilistic branching, deterministic, and assignment locations;
\item $\pvars=\{x_1,\dots,x_{|\pvars|}\}$ is a finite set of \emph{program 
variables} (note that $\pvars \subseteq \vars$) ;
\item $\locinit$ is an initial location and $\vecinit$ is an initial 
\emph{assignment vector};
% POKRACOVAT ODSUD
\item $\transitions\subseteq\locs\times\locs$ is a transition relation;
%whose members are tuples of
%the form $(\ell,i,\up,\ell')$, where $\ell$ and $\ell'$ are source and target
%program locations, respectively, $1\leq i \leq |\pvars|$ is a \emph{target 
%variable index}, and $u$ 
%is an 
%\emph{update element}, which can 
%be one of the following mathematical objects: %\begin{itemize}
%%\item 
%(a)~an affine function $u\colon \Rset^{|\pvars|}\rightarrow \Rset$;
%%\item 
%(b)~a distribution $d\in \mathcal{D}$; or
%%%\item 
%(c)~a set $R\subseteq \Rset$.
%\end{itemize}
\item $\updates$ is a function assigning to each transition outgoing from an 
assignment location a tuple $(i,\up)$, where $1\leq i \leq |\pvars|$ is a 
\emph{target 
	variable index} and $\up$ 
is an 
\emph{update element}, which can 
be one of the following mathematical objects: %\begin{itemize}
%\item 
(a)~a Borel-measurable function $u\colon \Rset^{|\pvars|}\rightarrow \Rset$;
%\item 
(b)~a distribution $d\in \mathcal{D}$; or
%\item 
(c)~a set $R\subseteq \Rset$.
%\end{itemize}
\item $\probdist=\{\prob_{\ell}\}_{\ell \in \locsPB}$ is a collection of
probability distributions, where each $\prob_{\ell}$ is a discrete probability
distribution on the set of all transitions outgoing from~$\ell$;
\item $\guards$ is a function assigning a propositionally linear predicate
(a \emph{guard}) over $\pvars$ to each transition outgoing from a deterministic 
location.
\end{compactitem}

@@

\smallskip\noindent{\em Executions and reachable configurations.}
We say that a configuration $(\loc',\vec{x}')$ is a \emph{successor} of a 
configuration $(\loc,\vec{x})$ if 
%A \emph{finite path} (or
%\emph{execution fragment}) of length $k$ in $\pCFG$ is a finite sequence of
%configurations $(\ell_0,\vec{x}_0)\cdots(\ell_k,\vec{x}_k)$ such that 
%%$(\loc_0,\vec{x}_0)=(\locinit,\vecinit)$ and 
%for each
%$0 \leq i < k$ 
there is a transition  $\tau=(\loc,\loc')$ enabled in
$(\loc,\vec{x})$ 
%such that $\vec{x}_{i+1}=\vec{x}_i(j\leftarrow a)$ where $a$ 
%satisfies one of the following:
and $\vec{x}'$ satisfies the following:
\begin{compactitem}
\item if $\loc$ is not an assignment location, then $\vec{x}'=\vec{x}$;
\item if $\loc$ is an assignment location with $\updates(\tau)=(j,\up)$, then 
$\vec{x}_{i+1}=\vec{x}_i(j\leftarrow a)$ where $a$ satisfies one of the 
following 
depending on the type of $\up$:
\begin{compactitem}
\item if $\up$ is a Borel-measurable function, then $a=\up(\vec{x})$;
\item if $\up$ is an integrable\footnote{A distribution on some numerical 
domain 
is integrable if its expected value exists and is finite. In particular, each 
Dirac distribution is integrable.} distribution $d$, then 
$a\in \support(d)$;
\item if $\up$ is a set, then $a$ is some element of $\up$.
\end{compactitem}
\end{compactitem}
A \emph{finite path} (or
\emph{execution fragment}) of length $k$ in $\pCFG$ is a finite sequence of
configurations $(\ell_0,\vec{x}_0)\cdots(\ell_k,\vec{x}_k)$ such that 
%$(\loc_0,\vec{x}_0)=(\locinit,\vecinit)$ and 
for each
$0 \leq i < k$ the configuration $(\loc_{i+1},\vec{x}_{i+1})$ is a successor of 
$(\loc_i,\vec{x}_i)$.
A \emph{run} (or \emph{execution}) in
$\pCFG$ is an infinite sequence of configurations whose every finite
prefix is a finite path.
A configuration $(\loc,\vec{x})$ is {\em reachable} from the initial 
configuration
$(\locinit,\vecinit)$
if there is a finite path starting in $(\locinit,\vecinit)$ that ends in
$(\loc,\vec{x})$. We denote by $\Conf_\pCFG,\Fpath_\pCFG$ and $\Run_\pCFG$ the 
sets of all configurations, finite paths and runs in $\pCFG$, respectively, 
dropping the index $\pCFG$ when known from the context.

@@

\smallskip\noindent{\em Termination, reachability, and termination time.}
In the following, consider a \PP{} $P$ and its associated pCFG $\pCFG_P$. 
We say that a run $\run$ of $\pCFG_P$ \emph{reaches} a set of configurations 
$\confset$ if it  contains a configuration from $\confset$. A run 
\emph{terminates} if it reaches a 
configuration whose first component (i.e. a location of $\pCFG_P$) is the 
location
$\loc_{P}^{\lout}$ corresponding to the value of the program counter after 
executing $P$. 
To each set of configurations $\confset$ we can assign a random variable 
$\treach{\confset}$ 
such that for each run $\run$ the value $\treach{\confset}(\run)$ represents 
the first 
point in time when 
the current configuration on $\run$ is in $\confset$. If a run $\run$ does 
\emph{not} 
reach a set $\confset$, then 
$\treach{\confset}(\run)=\infty$. We call $\treach{\confset}$ the 
\emph{reachability time} of $\confset$. In particular, if $\confset$ is the set 
of all configurations $(\loc,\vec{x})$ such that $\loc=\loc^\lout_P$ (the 
terminal location of $\pCFG_P$), then $\treach{\confset}$ is 
called a \emph{termination time}, as it returns the number of steps after which 
$\run$ terminates. Since termination time is an important concept on its own, 
we use a special notation $\ttime$ for it.
%\begin{example}
%Consider the program $P$ such that \hspace{-0.25cm}\newsavebox{\figa}
%\lstset{language=affprob}
%\lstset{tabsize=1}
%\begin{lrbox}{\figa}
%\begin{lstlisting}[mathescape]
%$P=$x:=2;y:=1
%\end{lstlisting}
%\end{lrbox}
%\usebox{\figa}. The associated pCFG has three locations: initial location 
%$\loc_{P}^{\lin}$, terminal location $\loc_{P}^{\lout}$, and one 
%``intermediate'' location, say $\loc$. In the pCFG consider a run 
%$\run=(\loc_{P}^{\lin},(0,0))(\loc,(2,0))(\loc_{P}^{\lout},(2,1))(\loc_{P}^{\lout},(2,1))\cdots$.
% Then $\ttime(\run)=2$. If we denote by $C$ the set of configurations 
% $\{(\loc',(x,y))\mid \loc' \in\{\loc_{P}^{\lin},\loc_{P}^{\lout},\loc\} \text{ 
% and } x\geq 2\}$, then $\treach{C}(\run)=1$.
%\end{example}
Since a probabilistic program may exhibit more than one run, we are interested 
in probabilities of runs that terminate or reach some set of configurations. 
This gives rise to the following fundamental computational problems regarding 
termination:
